A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-02 main.tex
4.4 - 5 edits

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\bG$ for $n$ subjects and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. In order to equitably compare SNP effects, regardless of allele frequency, in practice $\bG$ is standardized. We denote the standardized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 \pi_j) / \sqrt{\pi_j (1 - \pi_j)}$ and $\pi_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. Additionally, $\bX_0$ represents an $n \times p_0$ matrix of unpenalized covariates and the intercept. Thus, at a minimum, $\bX_0$ will be an $n \times 1$ vector of 1s to represent the intercept, in which case $\bbeta_0 \equiv \beta_0$ is the scalar intercept. However, $\bX_0$ may additionally include observed covariates such as age, sex, or smoking history, which are included in the model without penalty. We assume that the $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of these covariates, additive SNP effects, environmental effects, and observational noise, and consider the linear model,
\begin{equation}
    \label{eqn:model}
    \by = \bX_0 \bbeta_0 + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\bbeta_0$ is a $p_0 \times 1$ vector which includes the intercept term and the effects of any unpenalized covariates, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects among subjects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1.1]{figures/beta_mse.png}
    \caption{Estimation accuracy of different methods for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
    \label{fig:mse}
\end{figure}
\anna{talk about karl rohe stuff here - if anywhere}
In addition to MSE, these methods' performance in terms of variable selection accuracy were quantified using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the mean true positive rate and corresponding standard deviation in the same settings as those illustrated in Figure \ref{fig:mse}. The TPRs echo the MSE results in that Lasso and LMM-Lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-Lasso and PC-Lasso achieve substantial improvements in TPR over Lasso across data types. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Semi-Empirical data sets than in the 1D Linear Admixture data. Again, this is due to the reduced proportion of environmental confounding that is in the column space of $\boldsymbol{P}_{\bX}$, which results in a lesser degree of confounding of the SNP-phenotype associations.

@@

LMM-Lasso-ggmix achieves a moderate improvement in the estimation of $\eta$ compared to LMM-Lasso-Rakitsch, particularly in the presence of lower $\eta$ values and where $n << p$, as shown in the middle panel of Figure \ref{fig:eta_beta_mse}. These improved $\eta$ estimates do not correspond to smaller $\bbeta$ MSE for LMM-Lasso-ggmix. In addition, when $n \ge 1000$, LMM-Lasso-ggmix's estimation of $\eta$ suffers. When $n \ge 1000$ and in the presence of substantial environmental confounding, $\xi = 0.8$,  LMM-Lasso-ggmix's performance is hindered to the point that LMM-Lasso-Rakitsch estimates $\eta$ more accurately.

Overall these results show that LMM-Lasso-ggmix performs as well as LMM-Lasso-Rakitsch in estimating $\bbeta$ and outperforms LMM-Lasso-Rakitsch in estimating $\eta$ when $n << p$. However, these are also the settings in which LMM-Lasso-ggmix is most computationally inefficient. The bottom panel of Figure \ref{fig:eta_beta_mse} shows the computing time required of each method in the various simulation settings, and shows that when $n << p$, LMM-Lasso-ggmix is up to 60 times slower than LMM-Lasso-Rakitsch.

Conversely, in settings where $n \ge p$, LMM-Lasso-ggmix is up to 10 times faster than LMM-Lasso-Rakitsch. However, these are also settings in which LMM-Lasso Rakitsch outperforms LMM-Lasso-ggmix in estimating $\bbeta$ and $\eta$ accurately and consistently, likely due to the ill-conditioned RRM \citep{ledoit2004well}.

% Given the intended use of these methods for the analysis of genetic data, where the number of genetic variants often outnumber the number of subjects, these results could be of little consequence to practitioners.

@@

In this work, we have reviewed the concept of population stratification and explicitly defined the environmental mechanism by which it may confound the genotype-phenotype relationship. We have provided a broad outline of existing methods for ameliorating the effects of these unobserved confounding influences. In addition, we have provided a thorough review of two popular approaches for correcting for population structure in genetic data: PC-Lasso and LMM-Lasso.

@@

Unsurprisingly, we find that PC-Lasso and LMM-Lasso outperform the ordinary Lasso whenever environmental confounding effects are present. However, in the absence of such effects, PC-Lasso may hinder estimation accuracy and perform more poorly than even the ordinary Lasso. This is due to the fact that the principal components are unpenalized and thus included in the final model, even when they do more harm than good in explaining phenotypic variability. Additionally, we find that LMM-Lasso outperforms PC-Lasso when (1) fine subpopulation structure is present and (2) when confounding effects are nonlinear and discordant in nature. In the presence of coarse subpopulation structure which corresponds to concordant, dichotomous environmental effects, PC-Lasso may outperform LMM-Lasso in estimation accuracy. However, LMM-Lasso is more robust to subpopulation structure as well as nonlinear and discordant environmental effects. Therefore, LMM-Lasso is a more robust method for practitioners seeking to obtain unbiased estimates when the exact nature of the population structure and corresponding environmental heterogeneity are unknown.

Finally, we considered two implementations of the LMM-Lasso method: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although we found that both methods perform similarly in terms of $\bbeta$ estimation accuracy, LMM-Lasso-ggmix has several drawbacks. Firstly, it suffers from estimation instability in $n \ge p$ settings. Secondly, although LMM-Lasso-Rakitsch and LMM-Lasso-ggmix perform comparably when $n << p$, LMM-Lasso-ggmix requires up to 60 times the computing time of LMM-Lasso-Rakitsch in these settings due to its iterative estimation procedure. The additional time required of LMM-Lasso-ggmix may be worthwhile if the objective of an analysis is to accurately estimate the narrow-sense heritability in $n << p$ settings. However, the LMM-Lasso-Rakitsch method is more robust to varying levels of environmental confounding and sample sizes and thus considered by the authors to be more broadly applicable. \anna{talk about how ggmix seems to get stuck in a local minimum because it's esimating more parameters the optimization problem is not necessarily convex, this its prone to getting stuck. While this may make it fast, it makes it terrible for estimation accuracy. Reference iterations plots?}

@@

It is worth nothing that even in scenarios where LMM-Lasso outperforms other methods, its estimation accuracy still suffers due to the bias inherent in penalized regression methods. Because of this, bias-reduction techniques for penalized LMMs are an attractive avenue for future research.

\anna{the importance of distinguishing between population structure and heterogeneous environmental exposures in penalized regression methods - do we want to get into the importance of not conflating environmental effects with heritability?}

>>>>

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\bG$ for $n$ subjects and $p$ SNPs, where each $G_{ij} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. In order to equitably compare SNP effects, regardless of allele frequency, in practice $\bG$ is standardized. We denote the standardized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 \pi_j) / \sqrt{\pi_j (1 - \pi_j)}$ and $\pi_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. Additionally, $\bX_0$ represents an $n \times p_0$ matrix of unpenalized covariates and the intercept. Thus, at a minimum, $\bX_0$ will be an $n \times 1$ vector of 1s to represent the intercept, in which case $\bbeta_0 \equiv \beta_0$ is the scalar intercept. However, $\bX_0$ may additionally include observed covariates such as age, sex, or smoking history, which are included in the model without penalty. We assume that the $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of these covariates, additive SNP effects, environmental effects, and observational noise, and consider the linear model,
\begin{equation}
    \label{eqn:model}
    \by = \bX_0 \bbeta_0 + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\bbeta_0$ is a $p_0 \times 1$ vector which includes the intercept term and the effects of any unpenalized covariates, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects among subjects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1.1]{figures/beta_mse.png}
    \caption{Estimation accuracy of different methods for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
    \label{fig:mse}
\end{figure}
% \anna{talk about karl rohe stuff here - if anywhere}

In addition to MSE, these methods' performance in terms of variable selection accuracy were quantified using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the mean true positive rate and corresponding standard deviation in the same settings as those illustrated in Figure \ref{fig:mse}. The TPRs echo the MSE results in that Lasso and LMM-Lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-Lasso and PC-Lasso achieve substantial improvements in TPR over Lasso across data types. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Semi-Empirical data sets than in the 1D Linear Admixture data. Again, this is due to the reduced proportion of environmental confounding that is in the column space of $\boldsymbol{P}_{\bX}$, which results in a lesser degree of confounding of the SNP-phenotype associations.

@@

LMM-Lasso-ggmix achieves an improvement in the estimation of $\eta$ compared to LMM-Lasso-Rakitsch, particularly in the presence of lower $\eta$ values and where $n << p$, as shown in the middle panel of Figure \ref{fig:eta_beta_mse}. These improved $\eta$ estimates do not correspond to smaller $\bbeta$ MSE for LMM-Lasso-ggmix and thus do not seem to improve its estimation of the fixed genetic effects. In addition, when $n \ge 1000$, LMM-Lasso-ggmix's estimation of $\eta$ suffers. When $n \ge 1000$ and substantial environmental confounding is present, $\xi = 0.8$,  LMM-Lasso-ggmix's performance is hindered to the point that LMM-Lasso-Rakitsch estimates $\eta$ more accurately.

Overall these results show that LMM-Lasso-ggmix performs as well as LMM-Lasso-Rakitsch in estimating $\bbeta$ and outperforms LMM-Lasso-Rakitsch in estimating $\eta$ when $n << p$. However, these are also the settings in which LMM-Lasso-ggmix is most computationally inefficient. The bottom panel of Figure \ref{fig:eta_beta_mse} shows the computing time required of each method in the various simulation settings, and shows that when $n << p$, LMM-Lasso-ggmix is up to 60 times slower than LMM-Lasso-Rakitsch. Conversely, in settings where $n \ge p$, LMM-Lasso-ggmix is up to 10 times faster than LMM-Lasso-Rakitsch. However, these are also settings in which LMM-Lasso Rakitsch substantially outperforms LMM-Lasso-ggmix in estimating $\bbeta$ and $\eta$ accurately and consistently.

LMM-Lasso-ggmix's poor performance in these scenarios are likely attributable to a number of a factors such as an ill-conditioned RRM  in $n \ge p$ settings \citep{ledoit2004well}. In addition, because LMM-Lasso-ggmix attempts to estimate more parameters than LMM-Lasso-Rakitsch, its optimization problem is not necessarily convex, making it prone to getting stuck in local minima before it is able to extensively explore the coefficient path. An examination of the number of iterations required for convergence for each $\lambda$ in the coefficient path suggests that this is the case for our various simulation settings \anna{Supplementary figure ?}. Getting stuck in this manner would lead to few iterations required for each $\lambda$ value before convergence, allowing for an advantage in terms of computational speed, but poor estimation, as we see in Figure \ref{fig:eta_beta_mse}.

\anna{re-create niter\_parLambda fig focusing on the eta = xi = 0.8 seting and displaying all data points - not just averages}

@@

In this work, we have reviewed the concept of population stratification and explicitly defined an environmental mechanism by which it may confound the genotype-phenotype relationship. We have provided a broad outline of existing methods for ameliorating the effects of these unobserved confounding influences. In addition, we have provided a thorough review of two popular approaches for correcting for population structure in genetic data: PC-Lasso and LMM-Lasso.

@@

Unsurprisingly, we find that PC-Lasso and LMM-Lasso outperform the ordinary Lasso whenever environmental confounding effects are present. However, in the absence of such effects, PC-Lasso may hinder estimation accuracy and perform more poorly than even the ordinary Lasso. This is due to the fact that the principal components are unpenalized and thus included in the final model, even when they do more harm than good in explaining phenotypic variability. Additionally, we find that LMM-Lasso outperforms PC-Lasso in the presence of fine subpopulation structure and when confounding effects are nonlinear and discordant in nature. In the presence of coarse subpopulation structure which corresponds to concordant, dichotomous environmental effects, PC-Lasso may outperform LMM-Lasso in estimation accuracy. However, LMM-Lasso is more robust to subpopulation structure as well as nonlinear and discordant environmental effects. Therefore, LMM-Lasso is a more robust method for practitioners seeking to obtain unbiased estimates when the exact nature of the population structure and corresponding environmental heterogeneity are unknown.

We considered two implementations of the LMM-Lasso method: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although we found that both methods perform similarly in terms of $\bbeta$ estimation accuracy, LMM-Lasso-ggmix has several drawbacks. Firstly, it suffers from estimation instability in $n \ge p$ settings. Secondly, although LMM-Lasso-Rakitsch and LMM-Lasso-ggmix perform comparably when $n << p$, LMM-Lasso-ggmix requires up to 60 times the computing time of LMM-Lasso-Rakitsch in these settings due to its iterative estimation procedure. The additional time required of LMM-Lasso-ggmix may be worthwhile if the objective of an analysis is to accurately estimate the narrow-sense heritability in $n << p$ settings. However, the LMM-Lasso-Rakitsch method is more robust to varying levels of environmental confounding and sample sizes and thus considered by the authors to be more broadly applicable.

@@

As an additional point of interest, recall that we provide qualitative and mathematical explanations of how the degree to which differential environmental exposures may confound the genotype-phenotype relationship is dependent upon its similarity to population structure. While the objective of these explanations, and this work as a whole, is to elucidate the manner in which failing to correct for unobserved environmental effects can bias SNP effect estimates, it also highlights the manner in which such unobserved environmental effects have the potential to inflate residual model variance and bias narrow-sense heritability estimates. A more detailed treatment of narrow-sense heritability estimation is beyond the scope of this paper. However, we hope this emphasizes the broad implications and importance of the subtle distinction between the effect of population structure and unobserved environmental heterogeneity.

Finally, it is worth nothing that even in scenarios where LMM-Lasso outperforms other methods, its estimation accuracy still suffers due to the bias inherent in penalized regression methods. Because of this, bias-reduction techniques for penalized LMMs are an attractive avenue for future research.
